Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course

### 2025-03-12 paper/paper.qmd
addressed nathalie comments

---
title: "US President Prediction"
subtitle: "Predict the results of 2024 US President Election"
author: 
  - Yun Chu
  - Felix Li
  - Wen Han Zhao 
thanks: "Code and data are available at: [https://github.com/younazhao/US-President-Prediction/tree/main](https://github.com/younazhao/US-President-Prediction/tree/main)."
date: 22 October 2024
date-format: long
abstract: "In this study, we aim to study the 2024 U.S. Presidential Election with the methodology of  polls of polls across various states. With the insightful analyses on polls provided by Redfiled & Wilton Strategies, the conducted survey on the population highlights various key point for the prediction of US president. Using Generalized Linear Models (GLMs), this report analyzes the accuracy of prediction in president votes through several key variables from pollster. The model incorporates variables such as pollscore, methodology, transparency_score and hypothetical. The prediction finding highlights the significance of candicate trustworthiness among the voter decision, providing insights for swing voters. This analysis offers valuable information into the potential electoral outcomes driving the 2024 US president election. "
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(readr)
library(dplyr)
library(rstanarm)
library(bayesplot)
library(ggplot2)
library(ggpubr)
```

@@

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

## Overview

We use the statistical programming language R [@citeR].... Our data.... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

In this analysis, we aim to model the popularity trends for the two 2024 president nominates, Kamala Harris and Donald Trump, based on high-quality polling data collected at the national level after Harris’s declaration on July 21, 2024. We used Bayesian linear regression models with Gaussian error structures to estimate changes in polling percentages over time for each candidate.

## Data Filtering and Preparation

The data were first filtered to retain only high-quality polls, defined as those with a numeric grade of 2.0 or above and a transparency score of 4 or higher. We focused exclusively on polls where the *state* variable indicates national coverage, ensuring the poll represents nationwide opinions rather than a single state, which could introduce bias. Additionally, we filtered for polls conducted after July 21, 2024, the date when Kamala Harris announced her candidacy.

To prevent issues with missing data in our models, we excluded any rows where key variables (such as polling percentage or end date) were missing.

## Model Specification

Two separate Bayesian linear regression models were fitted using the @rs package. Both models specified the formula:

\begin{align}
Y_i | \mu_i,\sigma  & \sim \text{Normal}(\mu_i,\sigma) \\
\mu_i & = \beta_0 + \beta_1X_{i1}\\
\beta_0 & \sim \text{Normal}(50, 5), \\ 
\beta_1 & \sim \text{Normal}(0, 0.1)
\end{align}

Where $Y_i$ is the support rate for Harris or Trump, separately, and $X_{i1}$ is the number of days passed after the latest polling sample of Trump or Harris.

## Model Justification and Limitations
We opted to use two separate models for Trump and Harris, as their supporter demographics differ significantly in ways that are challenging to capture within a single or even a small set of variables @pew_harris_trump_2024. The polling dataset we obtained primarily reflects the characteristics and trends within individual polls rather than offering direct predictors of the presidential election outcome. To maximize the model’s reliability, we refined the dataset by selecting polls with high credibility rather than attempting to predict outcomes directly from the available data.

Our model uses a simple linear approach with time as the sole predictor variable. While this may introduce potential bias by making a simplifying assumption about the trend over time, it also has the advantage of being straightforward to fit and interpret. This trade-off allows us to capture general trends while minimizing the risk of overfitting to limited or potentially inconsistent data sources.

# Results {#set-result}
The results of the Bayesian linear regression models for Kamala Harris and Donald Trump are summarized in @tbl-model_summary_H and @tbl-model_summary_T.

```{r}
#| label: tbl-model_summary_H
#| echo: false
#| tbl-cap: "Summary statistics of the Bayesian model for Harris"
#| warning: false
# Load required packages
library(broom.mixed)  # For tidy model summaries from mixed models
library(dplyr)
library(readr)
library(knitr)  # For nicely formatted tables
library(modelsummary)

# Load the saved models
model1 <- readRDS(file = here::here("models/models_harris.rds"))

# Extract a tidy summary of the model
model_summary <- broom.mixed::tidy(model1, effects = "fixed", conf.int = TRUE) %>%
  select(term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Term = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `2.5% CI` = conf.low,
    `97.5% CI` = conf.high
  )

# Display the summary table
kable(model_summary, digits = 3)

```
```{r}
#| label: tbl-model_summary_T
#| echo: false
#| tbl-cap: "Summary statistics of the Bayesian model for Trump"
# Load required packages
library(broom.mixed)  # For tidy model summaries from mixed models
library(dplyr)
library(readr)
library(knitr)  # For nicely formatted tables

# Load the saved models
model2 <- readRDS(file = here::here("models/models_trump.rds"))

# Extract a tidy summary of the model
model_summary <- broom.mixed::tidy(model2, effects = "fixed", conf.int = TRUE) %>%
  select(term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Term = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `2.5% CI` = conf.low,
    `97.5% CI` = conf.high
  )

>>>>

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
  - Nathalie Moon
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is currently under review by the University of Toronto's Research Ethics Board. Contributions: RA: Conceptualization, Data Curation, Formal Analysis, Funding Acquisition, Investigation, Methodology, Project Administration, Software, Visualization, Writing – Original Draft Preparation, Writing – Review & Editing; LC: Formal Analysis, Visualization, Writing – Original Draft Preparation, Writing – Review & Editing; NM: Conceptualization, Methodology, Writing – Original Draft Preparation, Writing – Review & Editing. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest is similar to the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

```{r setup}
#| include: false

library(arrow)
library(here)
library(knitr)
library(tidyverse)
library(scales)
library(modelsummary)
library(bayesplot)

# Read in data
all_data <-
  read_parquet(here("data/analysis_data.parquet"))
```

@@

Trustworthy data science is the practice of conducting data analysis in a transparent, ethical, and reliable manner. These principles are upheld in various ways, through ongoing education, adherence to professional norms and, and implementation of reproducible workflows.  To stay current, data scientists must continually update their knowledge of best practices for transparency and reproducibility, foster a culture which values different perspectives and accountability, and critically consider the broader impact of data-driven decisions.  Recent advances, such as the wide release of user-friendly Large Language Models (LLMs), particularly OpenAI’s ChatGPT, have transformed the data science toolkit. These powerful tools for natural language processing and generation influence various aspects of data analysis and automation, further emphasizing the need for trustworthy practices.

Like all new tools LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into the mainstream conversation. LLMs have quickly been put to use in both industry and academia. By now, many people, especially educators and students, have some experience with LLMs, in both personal and professional contexts.

In the context of teaching statistics and data science, LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots to act as personalized tutors for students [@fulgencio2024developing; @afzal2024tailoring]. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity [@Eke2023], the development of critical thinking skills [@thiga2024generative], and what effective learning looks like [@BaidooAnu_2023].

In this paper we are interested in better understanding LLMs as a tool for producing trustworthy data science. We study how they were used by students in an upper-year undergraduate data science course and whether students who used LLMs tended to have higher scores than those who did not.

The tasks involved in a trustworthy data science workflow can be generally broken down into a number of key competencies [@Adhikari2021Interleaving; @Gibbs2021Building]. One is programming, which is done when cleaning, analyzing, and visualizing data often using programming languages like R or Python. Another is writing, which is primarily done when communicating results. The potential for LLMs to positively affect students' academic performance in data science can be clearly inferred from their already-demonstrated capability in those competencies in adjacent professional fields.

In terms of computer programming, @Peng_2023 found a positive impact of GitHub Copilot (an LLM-powered programming assistant) on productivity. Specifically, in an experiment involving 95 freelance programmers, they found that a treatment group of programmers with access to GitHub Copilot completed a standardized programming task 56% faster than the control group. Programmers with less experience saw the greatest improvements in productivity.

@DellAcqua_2023, focusing on management consulting tasks, provide evidence that LLMs can improve writing productivity. In a field experiment involving consultants from Boston Consulting Group they found that the use of OpenAI's GPT-4 led to a 25% increase in delivery speed of business tasks, most of which involved some writing, as well as a 40% increase in human-rated performance on those tasks. Similar to computer programming, these productivity increases were most pronounced for those with below average performance, with their output increasing by 43%.

On the other hand, @Valenzuela_2024 argue that LLMs lead to a loss of serendipity which may lead to less original work, and potentially de-skilling primarily with respect to programming ability, among other consequences. These outcomes could negatively affect students' effective learning of data science.

@Ellis_Slade_2023 take a more optimistic perspective and argue that LLMs are just another technology that will impact statistics and data science education. Similarly, @Tu_2024 acknowledge that LLMs can streamline many parts of a data science workflow. With that in mind, they suggest that data-scientists-in-training should shift their self-perspectives from primarily being an "analyst" to primarily being a "product manager" responsible for strategic oversight of the analysis carried out by LLMs.

At the high school level @Lazar_2023 conducted an informal survey of secondary school teachers and students on their opinions of ChatGPT, and found that while LLMs could: help creativity, provide academic support when teachers were unavailable, and model certain types of writing well. Teachers were also concerned about the potential of LLMs to limit students' learning in certain ways through over-reliance. Beyond academic integrity concerns, @Lazar_2023 found that teachers had similar concerns to @Valenzuela_2024 about de-skilling and an overall loss of agency in writing and critical thinking.

@Cahill_2024 surveyed undergraduate political science students on their attitudes toward, and usage of, AI tools. They found that the use of ChatGPT was widespread. However, they also found that many students lacked confidence in using AI for academic purposes. In particular, only 11% "strongly agreed" that they know how to use AI to improve their writing. Students had nuanced views on appropriate AI use. Many respondents felt that using it to write whole papers was inappropriate, but using it for basic tasks like general assistance, writing feedback and basic data visualization appropriate.

To understand the current state of LLMs as a tool for trustworthy data science, this paper focuses on the association between student academic performance and their LLM usage. Specifically, we examine the relationship between students' grades and self-reported measures of student LLM usage, as well as student attitudes toward LLMs in general. This is based on students' final papers and a survey, conducted in a third-year undergraduate data science course at the University of Toronto. By examining how students interact with and perceive LLMs as tools, and how these variables translate into student outcomes, current practice with regard to LLM integration in data science can be better understood, leading to better recommendations for their future development.

The remainder of this paper is structured as follows: @sec-data visualizes and analyzes survey data and coursework from students. @sec-model specifies a model used to investigate the relationship. @sec-results describes and analyzes the model's results. @sec-discussion discusses the implications of the findings for data science education and future research and practice at the intersection of LLMs and trustworthy data science.

@@

## Background

To investigate students' usage and attitudes towards LLMs and how they related to their academic performance, a dataset containing their usage and attitudes, coursework, and academic performance was constructed. This was based on three components:

1.  an optional survey;
2.  self-reported LLM usage; and
3.  student marks on their final paper.

All data are from the cohort of students taking STA302 "Methods of Data Analysis I" in the Winter 2024 semester at the University of Toronto. This course had 275 students initially enrolled which, reflecting a normal rate of attrition for undergraduate statistics courses at the University of Toronto, reduced to 154 students by the end of the semester. Assessment was heavily based on three papers submitted over the course of the 12 week semester.

The student marks that we analyze are based only on the final paper, which is done individually. By this stage, uninterested students have typically dropped the course, and students are familiar with course expectations. A typical paper submission is 10-20 pages, and requires students to conduct original research to answer a research question of interest to them. It reflects the skills typically used by a professional data scientist. Students are expected to develop a research question of interest to them, identify or collect data to answer the question, conduct statistical analysis, and write a short paper. Examples of final papers (shared with consent) include: @hannahyu; @emilysu; and @bennyrochwerg.

By the time they are working on their final paper, students have submitted and received feedback on two previous papers with similar requirements and rubrics to that of the final paper. Each paper has the same basic structure and expectations. Before the final paper is due, students have received feedback on all their previous work in the class (including their past papers) and there is an optional two-day period of peer review.

The pre-requisites of this course mean that the typical student is an upper-year undergraduate. Coding and writing are major parts of the course. Students are welcome to use R or Python, but the majority code in R because that is the programming language currently mostly taught in pre-requisite courses. All writing must be in English. The primary motivation for having students write three papers as the main assessment for the course is to give them the opportunity to create a public portfolio of work they can use to apply for jobs.

Throughout the semester students were encouraged to use LLMs. Formal instruction was provided twice during the semester. The first was a masterclass taught by a computer science faculty member on the ethics of using LLMs (see @Horton2024 for details). The second was a masterclass taught by a TA on writing with LLMs.

Data was collected from students through an optional end-of-course survey. [Appendix -@sec-survey-details] details the questions asked in the survey. Whether or not they consented to their data being used, all respondents received a 1% increase in their final course grade for their participation. Consenting responses were then matched to their final paper mark, as well as the GitHub repository for their final paper. The responses were anonymized by removing any personal references to the students themselves including, names, emails, student numbers, and GitHub links.

Data cleaning and analysis was done using the `R` statistical programming language [@citeR], and the `tidyverse` [@tidyverse], `janitor` [@janitor], `reshape2` [@reshape], and `readxl` [@readxl] packages.

## Survey data

There were 146 responses to the survey. Of these, 119 respondents provided authorization for their data to be collected and used. Four of those respondents submitted the survey twice, and after removing their second response, 115 responses remained. Of those, 15 respondents did not include a statement on LLM usage in the README of the GitHub repository of their final paper, leaving 100 responses. Finally 7 of those respondents did not provide a usable GPA response. This leaves 93 respondents that were of use and were merged based on student name.

All but one respondent completed the survey within 45 minutes (@fig-response-time-and-gpa-1). That one respondent took more than 4,000 minutes to complete the survey, suggesting they took a break while filling it out. That respondent is included in the analysis dataset, but without them, the average time to complete the survey was 11 minutes, and the standard deviation was 8 minutes.

```{r fig-response-time-and-gpa}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of respondents' survey response times and self-reported GPA
#| fig-subcap: ["Distribution of survey response times","\"What is your GPA?\""]
#| layout-ncol: 1

all_data |>
  # summarize(mean = mean(mins_complete/60),
  # standard_dev = sd(mins_complete/60))
  ggplot(aes(x = mins_complete/60)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Time to complete survey (minutes)", y = "Frequency") +
  theme_minimal()

all_data |>
  # summarize(mean = mean(what_is_your_gpa, na.rm = TRUE),
  #           standard_dev = sd(what_is_your_gpa, na.rm = TRUE))
  ggplot(aes(x = what_is_your_gpa)) +
  geom_histogram(binwidth = 0.1) +
  labs(x = "Self-reported GPA", y = "Frequency") +
  theme_minimal()
```

There is a wide distribution of self-reported GPAs (@fig-response-time-and-gpa-2). The majority of responses cluster around a B (3.0 out of 4.0), and the average is 3.06 with a standard deviation of 0.55. One factor that may affect the range is that the course is required for programs in the Statistics, Mathematics and Computer Science Departments. Self-reported GPAs also introduce the possibility of reporting bias. For instance, respondents may have provided their cumulative GPA, their most recent term's GPA, or could have misreported it entirely.

Students from a range of years took the course, however the majority of respondents were in their 3rd or 4th year of study (@tbl-year-dist). The course's prerequisite two-course sequence is typically completed by students in their second year, which makes it difficult for students to take this course earlier than their 3rd year.

```{r tbl-year-dist}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: \"What year are you?\"

all_data |>
  count(what_year_are_you) |>
  pivot_wider(
    names_from = what_year_are_you,
    values_from = n,
    values_fill = list(n = 0)
  ) |>
  kable()
```

Respondents had a varied self-perception of their coding and writing abilities (@fig-selfperception). Most respondents believe it is important to be good at writing, but many are either indifferent or do not like to write (@fig-selfperception-1). Respondents also do not find writing to be particularly easy, which could be associated with the reported relative antipathy toward writing. Most respondents were at least somewhat confident in their own writing abilities, but a substantial number felt otherwise. Although few respondents felt that they were confident in their writing ability, more felt that they were able to catch their mistakes, which could indicate a disconnect between how respondents perceive their work and how the work was evaluated.

```{r fig-selfperception}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Self-perception of coding and writing abilities
#| fig-subcap: ["\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\"", "\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\""]

# Writing
writing_perceptions_long <-
  all_data |>
  select(
    writing_is_easy_for_me,
    i_like_to_write,
    i_believe_it_is_important_to_be_a_good_writer,
    when_i_edit_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_writing,
    i_am_confident_in_my_overall_writing_ability
  ) |>
  pivot_longer(cols = everything(),
               names_to = "statements",
               values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "writing_is_easy_for_me" ~ "\"Writing is easy for me.\"",
      statements == "i_like_to_write" ~ "\"I like to write.\"",
      statements == "i_believe_it_is_important_to_be_a_good_writer" ~ "\"I believe it is important\nto be a good writer.\"",
      statements == "when_i_edit_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I edit it is easy for\nme to catch my mistakes.\"",
      statements == "i_feel_confident_sharing_my_writing" ~ "\"I feel confident\nsharing my writing.\"",
      statements == "i_am_confident_in_my_overall_writing_ability" ~ "\"I am confident in my\noverall writing ability.\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different\nfrom me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

# so that "Very different from me" is Red - better visuals
set1_palette <- c("#4DAF4A", "#377EB8", "#E41A1C")

writing_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(x = "Response", y = "Number of repondents", fill = "Response:") +
  scale_fill_manual(values = set1_palette) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")

# Coding
coding_perceptions_long <-
  all_data |>
  select(
    coding_is_easy_for_me,
    i_like_to_code,
    i_believe_it_is_important_to_be_a_good_coder,
    when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_code,
    i_am_confident_in_my_overall_coding_ability
  ) |>
  pivot_longer(cols = everything(),
               names_to = "statements",
               values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "coding_is_easy_for_me" ~ "\"Coding is easy for me\"",
      statements == "i_like_to_code" ~ "\"I like to code\"",
      statements == "i_believe_it_is_important_to_be_a_good_coder" ~ "\"I believe it is important\n to be a good coder\"",
      statements == "when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I check my code, it is\n easy for me to catch my\nmistakes\"",
      statements == "i_feel_confident_sharing_my_code" ~ "\"I feel confident\n sharing my code\"",
      statements == "i_am_confident_in_my_overall_coding_ability" ~ "\"I am confident in\n my overall coding ability\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different\nfrom me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

coding_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(x = "Response", y = "Number of repondents", fill = "Response:") +
  scale_fill_manual(values = set1_palette) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")
```

Respondent self-perceptions regarding coding proficiency and importance varied (@fig-selfperception-2). There is strong consensus on the perceived importance of coding skills. However, respondents' self-assessed coding abilities and enjoyment are more heterogeneous, with a substantial proportion reporting moderate rather than high levels of ease and enjoyment in coding tasks.

Overall there was a moderate level of confidence among respondents in their overall coding ability, willingness to share code, and capacity to identify errors (@fig-selfperception-2). Notably, respondents express slightly higher confidence in detecting their own coding mistakes compared to general coding ability or code sharing. These patterns suggest that while respondents have developed some coding self-efficacy, there is still considerable potential for enhancing their perceived competence and comfort across various coding-related activities.

The majority of respondents were at least "somewhat familiar" with generative AI such as ChatGPT (@tbl-ai-familiarity-usage-1). Though respondents' self-perceptions around writing and coding are varied, there was strong consensus that the use of generative AI tools is appropriate within an academic setting (@tbl-ai-familiarity-usage-2). Most respondents who responded "It depends" generally found artificial intelligence tools to be appropriate, though with certain guidelines and rules governing their use.

```{r tbl-ai-familiarity-usage}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Familiarity with, and appropriateness of, generative AI
#| tbl-subcap: ["'How familiar are you with using generative AI tools such as OpenAI's ChatGPT or equivalents?'","'To what extent do you think using generative AI tools such as ChatGPT by OpenAI (or equivalents) is ethical and appropriate for schoolwork?'"]

all_data |>
  count(ai_familiarity) |>
  kable(col.names = c("Extent of AI familiarity", "Students"))

all_data |>
  count(ai_schoolwork_appropriate) |>
  kable(col.names = c("Ethical & Appropriate for school?", "Students"))
```

To understand the role of LLMs in learning, students identified their usage in a more granular way by selecting various pre-defined use cases in the survey (@fig-AI-use-and-usefulness-1). Technical questions and explaining concepts were the two top use cases among students in the course. More than half of students also used LLMs for quick questions, general knowledge, writing paper code, and checking solutions. Just less than half used it to write paper content, which could suggest that students do not feel confident using it to improve writing; this could be similar to the attitudes expressed by students surveyed in @Cahill_2024.

```{r fig-AI-use-and-usefulness}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: "Use and usefulness of generative AI"
#| fig-subcap: ["\"If you have used generative AI tools such as OpenAI's ChatGPT or equivalents, in what ways have you used it (select all that apply)?\"", "\"How helpful did you find generative AI tools such as ChatGPT by OpenAI (or equivalents) for each component of STA302?\""]

# Use
all_data |>
  select(how_have_you_used) |>
  separate_rows(how_have_you_used, sep = ";") |>
  mutate(
    how_have_you_used = case_when(
      how_have_you_used == "check grammar mistake" ~ "Other",
      how_have_you_used == "debugging code" ~ "Other",
      how_have_you_used == "mental health support" ~ "Other",
      how_have_you_used == "Making practice questions for me with increasing difficulty" ~ "Other",
      how_have_you_used == "Generating test questions" ~ "Other",
      TRUE ~ how_have_you_used
    )
  ) |> 
  count(how_have_you_used) |>
  filter(how_have_you_used != "") |>
  arrange(desc(n)) |>
  ggplot(aes(x = n, y = reorder(how_have_you_used, n))) +
  geom_bar(stat = "identity") +
  labs(x = "Number of respondents", y = "Type of usage") +
  theme_minimal()

# Usefulness
all_data |>
  select(
    weekly_quiz,
    weekly_mini_essay,
    papers_generating_ideas,
    papers_writing_code,
    papers_writing_content,
    papers_improving_content
  ) |>
  pivot_longer(cols = everything(),
               names_to = "task",
               values_to = "response") |>
  mutate(
    # Simplify response categories and rename tasks
    response_simplified = case_when(
      response %in% c(
        "\"Not helpful\"",
        "\"I did not use generative AI for this component\""
      ) ~ "Less Helpful",
      response %in% c("\"Somewhat helpful\"", "\"Very helpful\"") ~ "More Helpful",
      TRUE ~ "Unknown"  # Handle unexpected values
    ),
    task = case_when(
      task == "weekly_quiz" ~ "Weekly Quiz",
      task == "weekly_mini_essay" ~ "Weekly Mini Essay",
      task == "papers_generating_ideas" ~ "Generating Ideas",
      task == "papers_writing_code" ~ "Writing Code",
      task == "papers_writing_content" ~ "Writing Content",
      task == "papers_improving_content" ~ "Improving Content",
      TRUE ~ "Other"  # Handle any unexpected task names
    )
  ) |>
  group_by(task, response_simplified) |>
  summarise(count = n(), .groups = 'drop') |>
  ggplot(aes(x = response_simplified, y = count, fill = response_simplified)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ task, scales = "free_y") +
  labs(y = "Number of respondents", x = "", fill = "Response type: ") +
  theme_minimal() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(axis.text.y = element_blank(), legend.position = "bottom")
```

Respondents were also asked to rate the helpfulness of LLMs on various tasks assigned during the course on a 4-point scale of "Did not use" to "Very Helpful" (@fig-AI-use-and-usefulness-2). To simplify the presentation, responses were grouped into two main categories: "Less Helpful" and "More Helpful." The "Less Helpful" category combines responses where students found the AI either "Not helpful" or did not use it for the task, while the "More Helpful" category includes responses where the LLM was considered "Somewhat helpful" or "Very helpful".

Respondents differed in terms of how they used LLMs in the course (@fig-AI-use-and-usefulness-2). While most tasks were roughly split between respondents finding LLMs helpful or not, respondents found them most helpful in generating code. In the context of the course, this mostly meant generating R code for transforming, analyzing, and visualizing data. To a lesser extent, respondents also found LLMs to be helpful in improving the existing writing they had, while at the same time not favouring it for writing content from scratch.

Finally, one question asked students to elaborate on whether they thought generative AI tools such as ChatGPT were ethical and appropriate for schoolwork. This was an open-response question. To provide a sense of the responses, we used Anthropic's Claude 3.5 Sonnet model (as at 5 August 2024) to summarize the comments and it provided:

> Many students view AI as a helpful supplementary tool, comparing it to resources like Google or calculators. They believe it can aid in understanding concepts, debugging code, brainstorming ideas, and saving time on routine tasks. However, there's a consensus that AI should not be used to complete entire assignments or replace original thinking. Students emphasize the importance of using AI ethically, citing it when appropriate, and not relying on it exclusively. Some argue that learning to use AI effectively is a valuable skill for future careers. Concerns raised include the potential for plagiarism, the risk of hindering critical thinking skills, and the possibility of receiving incorrect information. Overall, most students support the responsible use of AI in education, with proper guidelines and transparency, while recognizing the need to maintain academic integrity and develop independent learning skills.

## LLM usage and final paper marks

Two other components were merged with the survey responses: self-reported LLM usage on the final paper, and final paper mark.

Students were encouraged to use LLMs to complete their papers. Each paper required the students to disclose their usage through a statement in the GitHub repository README for the paper. Even students who did not use generative AI at all were required to state this in the README. For students who did use generative AI, there was an additional requirement, where possible, that they save the logs of their usage in a txt file which was also included in their GitHub repository.

Those README statements were gathered and parsed using OpenAI's ChatGPT 4o model (as at 26 July). The following prompt was used:

> The following statement is about to what extent LLMs were used by a student. Please characterize it as one of: "None", "Minimal", "Somewhat", "Extensive", "Unsure". Respond with only one of those options.

All classifications were then manually checked for reasonableness.

We find a varied extent of self-reported LLM usage (@tbl-llm-usage). 41 respondents were classified as having made extensive use of LLMs, while 28 were classified as having made somewhat use. 31 respondents were classified as having made minimal or no use of LLMs. In the analysis dataset we combine those two classifications because only 8 respondents were classified as having minimal usage.

```{r tbl-llm-usage}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Self-reported LLM usage in final paper

all_data |>
  count(llm_usage) |> 
  kable(col.names = c("Self-reported LLM usage", "Number"))
```

The third, and final, component is the mark, in percentages, on the final paper (@fig-mark-1). The overall mean was 78% and standard deviation was 17 percentage points. However there was considerable differences by the extent of LLM usage (@tbl-mark; @fig-mark-2).

```{r fig-mark}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of marks on final paper (in percentages)
#| fig-subcap: ["Overall", "By self-reported LLM usage"]

all_data |>
  select(mark) |>
  ggplot(aes(x = mark)) +
  geom_histogram(binwidth = 0.05) +
  scale_x_continuous(labels = label_percent()) +
  labs(x = "Final paper mark (percentage)", y = "Count") +
  theme_minimal()

all_data |>
  ggplot(mapping = aes(x = mark)) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(labels = label_percent()) +
  labs(x = "Final paper mark (percentage)", y = "Count") +
  facet_wrap(vars(llm_usage),
             ncol = 2,
             labeller = labeller(
               llm_usage = c(
                 "Extensive" = "Extensive LLM usage",
                 "Minimal" = "Minimal LLM usage",
                 "None" = "No LLM usage",
                 "Somewhat" = "Some LLM usage"
               )
             )) +
  theme_minimal()
```

```{r tbl-mark}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Self-reported LLM usage in final paper and final paper mark

all_data |>
  group_by(llm_usage) |>
  summarize(mean = mean(mark) |> round(2),
            sd = sd(mark) |> round(2)) |>
  kable(col.names = c("Self-reported LLM usage", "Mean", "Std dev"))
```

@@

The goal of our modelling strategy is to better understand how a respondents' result on their final paper associates with their self-reported LLM usage. Students received their result on the final paper as a proportion (a 0-100% mark), so we use a Bayesian zero-one-inflated beta regression to best fit the distribution in this interval. We use this type of regression given the ability of the beta distribution to effectively describe continuous and skewed  (heteroskedastic) datasets, which we can see in the asymmetrical histogram of the distribution of marks.

Transforming the regression with a logit function as the link function ensures that the predicted values stay between zero and one, respecting the boundaries of the data.

We briefly describe the model that we use, which follows @kurztotherescue, which was adapted from @ospina. No students got zeroes, though some got full marks (100%) on the paper, so we adapt the regression to one-inflated beta model as described by @heisstotherescue to account for this.

Model diagnostics are included in [Appendix -@sec-model-details].

The model uses a one-inflated beta regression approach to analyze the relationship between student LLM usage, GPA, and final paper marks. Let $Y_i$ represent the mark for student $i$, modeled as:

$$Y_i \sim \text{Zero-One-Inflated Beta}(\mu_i, \phi_i, \pi_{zoi_i}, \pi_{coi_i})$$

where:

- $\mu_i$ is the mean of the beta distribution (location parameter)
- $\phi_i$ is the precision parameter (controlling dispersion)
- $\pi_{zoi_i}$ is the probability of $Y_i$ being exactly 0 or 1
- $\pi_{coi_i}$ is the probability of $Y_i$ being in the continuous part of the distribution

The model components are specified as:

1. Mean component:
   $logit(\mu_i) = \beta_{0,\mu} + \beta_{1,\mu} \cdot \text{llm\_usage}_i + \beta_{2,\mu} \cdot \text{GPA}_i$

2. Precision component:
   $\log(\phi_i) = \beta_{0,\phi} + \beta_{1,\phi} \cdot \text{llm\_usage}_i + \beta_{2,\phi} \cdot \text{GPA}_i$

3. Zero-one inflation component:
   $logit(\pi_{zoi_i}) = \beta_{0,zoi} + \beta_{1,zoi} \cdot \text{llm\_usage}_i + \beta_{2,zoi} \cdot \text{GPA}_i$

4. Continuous outcome inflation component:
   $logit(\pi_{coi_i}) = \beta_{0,coi}$

The "llm_usage" predictor is categorical, with students' self-reported LLM usage classified as "Extensive" (reference level), "Somewhat", or "None or minimal". GPA is a continuous variable representing students' self-reported grade point averages. The model is implemented using the brms package in R, which provides a Bayesian framework for fitting these regression models.

The expected relationship between LLM usage and final mark is unclear, given the modelling strategy and variables collected. It could be that stronger students used LLMs in a more sophisticated way or that they did not need to use them. However, we can expect the relationship between GPA and final mark to be positive.

# Results {#sec-results}

Our key results are summarized in @tbl-model-results and @fig-model-results, with full output available in the Appendix. We draw on the `modelsummary` [@modelsummary] package to summarize the results.

@tbl-model-results and @fig-model-results present the key results from two models estimating the association between LLM usage, GPA and final paper mark. The first model includes only LLM usage, while the second model also includes self-reported GPA as a predictor.

```{r tbl-model-results}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Selected Coefficient estimates and mean absolute deviation (MAD)

fit1 <-
  readRDS(file = here("models/fit_one_inflated.rds"))

fit2 <-
  readRDS(file = here("models/fit_one_inflated_gpa.rds"))

modelsummary(
  list(
    "Base model" = fit1,
    "Self-reported GPA Included" = fit2
  ),
  statistic = "mad",
  fmt = 2,
  output = "kableExtra",
  kable_options = list(
    booktabs = TRUE,
    linesep = ""
  ),
  coef_map = c(
    "b_llm_usageSomewhat" = "LLM usage: Somewhat",
    "b_llm_usageNoneorminimal" = "LLM usage: None or minimal",
    "b_what_is_your_gpa" = "GPA", 
    "b_zoi_Intercept" = "Zero-one inflation intercept"
  )
)
```
